A Mechanised Library for Generic (Co)Programming

### 2024-02-06 paper/paper.tex
Containers

In Coq, to construct least/greatest fixpoints of functors, these need to build
\emph{strictly positive types}. A common approach to construct such strictly
positive endofunctors is to use \emph{container functors}~\cite{Whatever}.
However, reasoning about container and container functor equality requires to
consider: (a) functional extensionality; (b) heterogeneous equality. This is
problematic, due to our goals to remain axiom-free. Furthermore, standard
representations of containers in Coq will lead to OCaml code that contains
\mintinline{OCaml}{Obj.magic}.

A containers is defined by a type of \emph{shapes} (\mintinline{Coq}{Shape }),
and a \emph{family of position types} indexed by shapes
(\mintinline{Coq}{Pos}).  An \emph{extension} of this container is a functor,
whose action on objects is given by the following definition:
\begin{minted}{Coq}
  Record Container (Shape : Set) (Pos : Shape -> Set) (X : Set) :=
    { shape : Shape; contents : Pos shape -> X }.
\end{minted}
Note that, in our mechanisation, we call container extensions
\mintinline{Coq}{Container}. Generally, any functor that is isomorphic to a
container extension is called a \emph{container functor}. However, in our work
we will call \emph{container functor} to the functors defined by a specific
container extension.

The first problem in extracting Coq code from such container extensions is
the use of families of types. Since OCaml's

\dcas{A better approach for extraction. Unfortunately, to remain axiom-free and
heterogeneous equality we need to have a decidable ``Valid''}
\begin{minted}{Coq}
  Record Container (Shape : Set) (Pos : Set) 
                   (Valid : Shape -> Pos -> Prop) (X : Set) :=
    { shape : Shape; contents : {p | Valid shape pos} -> X }.
\end{minted}

>>>>

In Coq, to construct least/greatest fixpoints of functors, these need to build
\emph{strictly positive types}. A common approach to construct such strictly
positive endofunctors is to use \emph{container functors}~\cite{Whatever}.
However, reasoning about container and container functor equality requires to
consider: (a) functional extensionality; (b) heterogeneous equality. This is
problematic, due to our goals to remain axiom-free. Furthermore, standard
representations of containers in Coq will lead to OCaml code that contains
\mintinline{OCaml}{Obj.magic}. We will explain why, and our mechanisation of
containers in the remainder of this section.

A container is defined by a type of \emph{shapes} (\mintinline{Coq}{Shape }),
and a \emph{family of position types} indexed by shapes
(\mintinline{Coq}{Pos}).  An \emph{extension} of this container is a functor,
whose action on objects is given by the following definition:
\begin{minted}{Coq}
  Record Container (Shape : Set) (Pos : Shape -> Set) (X : Set) :=
    { shape : Shape; contents : Pos shape -> X }.
\end{minted}
In our mechanisation, we use \mintinline{Coq}{Container} to represent container
extensions.  Generally, any functor that is isomorphic to a container extension
is called a \emph{container functor}. However, in our work we will call
\emph{container functor} to the functors defined by a specific container
extension.

The first problem in extracting Coq code from such container extensions is the
use of families of types. Since OCaml's type system cannot accurately capture
such families of types, Coq's code extraction mechanism will insert unsafe
casts. To avoid unsafe casts, we can consider representing the positions of a
container as a type that captures \emph{all} the potential positions in all
possible shapes, together with a predicate \mintinline{Coq}{Valid} that
determines whether a position is valid in a shape. The family of positions is
then equivalent to the type \mintinline{Coq}{fun shape => {p | Valid shape p}}.
\begin{minted}{Coq}
  Record Container (Shape : Set) (Pos : Set) 
                   (Valid : Shape -> Pos -> Prop) (X : Set) :=
    { shape : Shape; contents : {p | Valid shape p} -> X }.
\end{minted}
This approach solves the the problem of the unsafe casts, since the code
extracted for \mintinine{Coq}{{p | Valid shape pos}} will be an OCaml
singleton type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of shapes
and positions.,
